---
title: "Metaprogram evaluation"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    code_folding: hide
params:
  metaprograms_object_file: NULL
  metaprograms_metrics_file: NULL
  ora_results_file: NULL
  combined_scores_file: NULL
  seed: 2026
---

This notebook evaluates a single set of metaprograms for a group of samples.
Metaprograms should be:

* Non-redundant
* Composed of a mix of libraries
* Biologically interpretable

The metrics shown here will help evaluate if the identified MPs show these characteristics.

```{r define_functions}
#| warning: FALSE

# helper function for getting jaccard between two vectors
calc_jaccard <- function(vec1, vec2) {
  length(intersect(vec1, vec2)) / length(union(vec1, vec2))
}

# function for getting the jaccard index between two gene sets
geneset_jaccard <- function(geneset1, geneset2, geneset_map = all_term_df) {

  genes1 <- geneset_map |>
    dplyr::filter(gs_name == geneset1) |>
    dplyr::pull(ensembl_gene)

  genes2 <- geneset_map |>
    dplyr::filter(gs_name == geneset2) |>
    dplyr::pull(ensembl_gene)

  jaccard <- calc_jaccard(genes1, genes2)

  return(jaccard)
}

# make a heatmap looking at the Jaccard index betwen gene sets for each MP
ora_jaccard_heatmap <- function(results_df, title){
  # only one list here since we will compare each MP to all other MPs
  id1_list <- split(results_df$ID, results_df$metaprogram)
  # create the grid of comparisons
  cross_df <- tidyr::expand_grid(id1 = names(id1_list), id2 = names(id1_list))

  # calculate a single Jaccard index for each combination using split lists & ids
  jaccard_scores <- cross_df |>
    purrr::pmap_dbl(\(id1, id2){
      calc_jaccard(id1_list[[id1]], id1_list[[id2]])
    })

  # add scores to the comparison grid and convert to matrix
  jaccard_matrix <- cross_df |>
    dplyr::mutate(jaccard = jaccard_scores) |>
    # convert to matrix
    tidyr::pivot_wider(
      names_from = "id2",
      values_from = "jaccard"
    ) |>
    tibble::column_to_rownames(var = "id1") |>
    as.matrix()

  ComplexHeatmap::Heatmap(jaccard_matrix,
                          name = "Jaccard index",
                          column_title = title,
                          col = circlize::colorRamp2(c(0, max(jaccard_matrix)), colors = c("white", "darkslateblue")))
}

# make a dotplot for ORA results from a single metaprogram
# modified from what we get with enrichplot::dotplot to take as input a dataframe
dotplot <- function(df, title = NULL) {

  df <- df |>
    # only show the top 10 results
    dplyr::slice_min(order_by = p.adjust, n = 10, with_ties = FALSE) |>
    dplyr::mutate(
      # convert GeneRatio to a numeric
      top_number = stringr::word(GeneRatio, 1, sep = "/") |> as.numeric(),
      bottom_number = stringr::word(GeneRatio, -1, sep = "/") |> as.numeric(),
      GeneRatio = top_number/ bottom_number,
      # reformat the description shown for y labels
      # remove _, wrap and reorder by GeneRatio
      Description = stringr::str_replace_all(Description, "_", " ") |>
        stringr::str_wrap(width = 30) |>
        forcats::fct_reorder(GeneRatio)

      )

  ggplot2::ggplot(df, aes(
    x     = GeneRatio,
    y     = Description,
    color = p.adjust,
    size  = Count
  )) +
    geom_point() +
    scale_color_continuous(
      low = "#e06663", high = "#327eba",
      name = "p.adjust",
      guide = guide_colorbar(reverse = TRUE)
    ) +
    scale_size(range = c(2, 10)) +
    labs(title = title, x = "GeneRatio", y = NULL)
}
```


```{r setup, message = FALSE, echo = FALSE}
library(ggplot2)

# Set default ggplot theme
theme_set(
  theme_bw() +
    theme(
      strip.background = element_rect(fill = "transparent", linewidth = 0.5)
    )
)

# some helpful settings
options(
  dplyr.summarise.inform = FALSE,
  DT.options = list(
    scrollX = TRUE,
    scrollY = "400px",
    scrollCollapse = TRUE,
    paging = FALSE,
    searching = FALSE
  )
)

set.seed(params$seed)
```


```{r}
# check for files to make sure they exist
stopifnot(
  "Metaprogram object file does not exist" = file.exists(params$metaprograms_object_file),
  "Metaprogram metrics file does not exist" = file.exists(params$metaprograms_metrics_file),
  "ORA results file does not exist" = file.exists(params$ora_results_file),
  "Metaprogram scores file does not exist" = file.exists(params$combined_scores_file)
)
```


```{r}
# read in objects
mp_obj <- readr::read_rds(params$metaprograms_object_file)

# read in metrics and ora results
mp_metrics_df <- readr::read_tsv(params$metaprograms_metrics_file)
ora_results_df <- readr::read_tsv(params$ora_results_file)

# pull out some items we are going to use multiple places to decrease typing later
cor_mtx <- mp_obj$cor_matrix
all_cor_df <- mp_obj$spectra_cor_df
```

```{r}
# read in tsv with all scores for all samples
score_df <- readr::read_tsv(params$combined_scores_file, col_types = list(.default = "c", mp_score = "d")) |>
  dplyr::mutate(
    sample_id = stringr::word(unique_id, -1, sep = "-")
  )
```


## Parameters

This notebook shows results for the following parameters:

* Number of metaprograms (k): `r mp_obj$n_metaprograms`
* Filtering of orphan spectra: `r mp_obj$filter_spectra`
* Oprhan spectra cutoff (max cross sample correlation): `r mp_obj$orphan_cutoff`
* Range of k used with cNMF (`all` indicates 5, 10, 15, 20, and 25 were used): `r mp_obj$cnmf_k_range`

Metaprograms were derived from:

* `r length(mp_obj$unique_ids)` libraries

```{r, results='asis'}
missing_ids <- mp_obj$removed_unique_ids
if(length(missing_ids) > 0){

  escaped_missing_ids <- htmltools::htmlEscape(missing_ids)
  missing_ids_bullets <- paste0("<li>", paste(escaped_missing_ids, collapse = "</li><li>"), "</li>")

    glue::glue("
      <div class=\"alert alert-info\">

      Metaprograms do not contain cNMF spectra from all libraries.
      All spectra from the following libraries were removed during filtering.

      {missing_ids_bullets}

      </div>
    ")
}
```


## Redundancy of metaprograms

Here we look at the correlation of all MPs to all other MPs.
Ideally, MPs are unique and non-redundant and therefore have low correlation to all other MPs.
The first plot shows all the spectra that comprise the MPs correlation to all other spectra and the second looks at correlation of each MP to all other MPs.

```{r}
# define colors for the cluster annotations
mp_colors <- palette.colors(n = length(mp_obj$cluster_list), palette = "alphabet") |>
  purrr::set_names(names(mp_obj$cluster_list))

# metaprogram/cluster annotation
annotation <- ComplexHeatmap::HeatmapAnnotation(
  metaprogram = sprintf("MP%02d", mp_obj$clusters),
  col = list(
    metaprogram = mp_colors
))


# get dendrogram
dist_mtx <- as.dist(1-cor_mtx)
tree <- hclust(dist_mtx, method = "ward.D2")
dend <- as.dendrogram(tree)

ComplexHeatmap::Heatmap(
  cor_mtx,
  show_row_names = FALSE,
  show_column_names = FALSE,
  cluster_rows = dend,
  cluster_columns = dend,
  top_annotation = annotation,
  name = "correlation"
)
```


```{r}
# combine all mps into a matrix for a heatmap
# get correlation to each other
mp_mtx <- do.call(cbind, mp_obj$metaprogram_list)
mp_cor_matrix <- cor(mp_mtx, method = "pearson", use = "pairwise.complete.obs")

annotation <- ComplexHeatmap::HeatmapAnnotation(
  metaprogram = colnames(mp_mtx),
  col = list(
    metaprogram = mp_colors
))

ComplexHeatmap::Heatmap(
  mp_cor_matrix,
  name = "Correlation",
  top_annotation = annotation
)
```


Next we will look at the correlation of each spectra to all other spectra within the same MP and across MPs.
Ideally, the spectra within a single MP should have high correlation to each other and low correlation to spectra from other MPs.
In the top left corner of each panel the difference in the median between the two groups (within and between MPs) is displayed.


```{r}
# convert the mp cluster list into a dataframe with one row per spectra
# use this to get the assigned mp for each cluster when calculating metrics
mp_lookup <- tibble::enframe(mp_obj$cluster_list, name = "metaprogram", value = "spectra") |>
  tidyr::unnest_longer(spectra)

# for each correlation, add in the MPs being compared and classify MP
mp_cor_df <- all_cor_df |>
  dplyr::filter(spectra1 != spectra2) |>
  dplyr::left_join(mp_lookup, by = c("spectra1" = "spectra")) |>
  dplyr::rename(mp1 = metaprogram) |>
  dplyr::left_join(mp_lookup, by = c("spectra2" = "spectra")) |>
  dplyr::rename(mp2 = metaprogram) |>
  dplyr::mutate(within_mp = mp1 == mp2) |>
  dplyr::mutate(mp_comparison_group = dplyr::if_else(within_mp, "Within MP", "Between MPs"),
                spectra_comparison_group = dplyr::case_when(
                  spectra1 == spectra2 ~ "self",
                  spectra1 < spectra2 ~ "different",
                  spectra1 > spectra2 ~ "different_dup"
                )) |>
  # only report comparisons between different spectra
  # and remove duplicates
  dplyr::filter(spectra_comparison_group == "different") |>
  dplyr::select(spectra1, spectra2, mp1, mp2, correlation, mp_comparison_group) |>
  dplyr::distinct()
```


```{r}
# grab the median difference
# `mp_cor_df` facets on `mp1`, so the label data needs the same column name to be placed on the
# matching panel rather than repeated on every panel
mp_labels <- mp_metrics_df |>
  dplyr::select(mp1 = metaprogram, median_difference) |>
  dplyr::mutate(x = -Inf, y = Inf)
```

```{r}
#| fig.width: 8

ggplot(mp_cor_df, aes(x = correlation, color = mp_comparison_group, fill = mp_comparison_group)) +
  geom_density(alpha = 0.25) +
  facet_wrap(vars(mp1), scales = "free_y") +
  scale_color_manual(values = c("Within MP" = "#00274C",
                                "Between MPs" = "#FFCB05")) +
  scale_fill_manual(values = c("Within MP" = "#00274C",
                               "Between MPs" = "#FFCB05")) +
  labs(x = "Correlation",
       y = "Density",
       fill = "MP comparison group",
       color = "MP comparison group") +
  geom_text(
    data = mp_labels,
    aes(x = x, y = y, label = round(median_difference, 2)),
    inherit.aes = FALSE,
    hjust = -0.1, vjust = 1.2,
    size = 4
  )
```


## Library diversity

<!--TODO: Update this to reflect samples and not just libraries-->

MPs should represent a mix of libraries.
This metric can be useful in filtering out any MPs that are sample-specific.

We will use Shannon's entropy to measure the diversity of each MP where a normalized value of 1 indicates high diversity.
Shannon's entropy is then used to calculate the normalized effective sample size or Hill number of order 1.
This represents the effective number of common or abundant samples.
The values are normalized by the total number of samples in the metaprogram such that values close to 1 indicate all samples are equally abundant.

```{r}
# grab shannons entropy and effective sample size from metrics
mp_metrics_df |>
  dplyr::select(
    metaprogram,
    # note that the names of the columns is samples, but this is really libraries right now
    num_samples,
    proportion_of_samples,
    shannon_entropy_norm,
    eff_number_norm
  ) |>
  DT::datatable(colnames = c("Metaprogram", "Num Libraries", "% Libraries", "Normalized Shannon's Entropy", "Normalized Effective Sample Size")
  ) |>
  DT::formatPercentage(c("proportion_of_samples")) |>
  DT::formatRound(c("shannon_entropy_norm", "eff_number_norm"))
```

The chart below shows the proportion of spectra from each metaprogram that are derived from individual libraries.
Each box represents a unique library, and its length represents that library's proportion of spectra contributing to the metaprogram.

```{r}
# add library Id column to metaprogram/spectra map
# TODO: Update this to use unique id to join and not sample id or library id
mp_lookup <- mp_lookup |>
  dplyr::mutate(
    # spectra are named with libraryid-sampleid_cnmf info
    sample_id = stringr::str_remove(spectra, "_k[0-9]+_CNMF[0-9]+$")
  )

total_libraries <- length(mp_obj$unique_id)

mp_prop_df <- mp_lookup |>
  dplyr::group_by(metaprogram, sample_id) |>
  # generate one row per mp/sample combination with the total count
  dplyr::summarise(n = dplyr::n(),
                   .groups = "drop_last") |>
  # get the frequency of each sample compared to all other samples
  dplyr::mutate(p = n / sum(n),
                num_samples = dplyr::n_distinct(sample_id)
  ) |>
  # ungroup before setting levels otherwise the plot doesn't work
  dplyr::ungroup() |>
  dplyr::mutate(metaprogram = forcats::fct_reorder(metaprogram, num_samples))

ggplot(mp_prop_df, aes(x = p, y = metaprogram, fill = sample_id)) +
  geom_col(color = "gray70") +
  labs(
    y = "Metaprogram",
    x = "Proportion of spectra",
    fill = "Sample"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```


## Biological interpretability

Ultimately, we want to be able to easily interpret the MPs.
We won't actually annotate them here, but we will look to see how biologically distinct the MPs might be.

To do that, we'll look at the results from running over-representation analysis on the top genes for each metaprogram.

```{r}
# pull out the top genes
mp_top_list <- mp_obj$top_genes
```

```{r}
#| results: 'asis'

# print out a message about which MPs have 0 genesets
all_mps <- names(mp_top_list)
missing_mps <- all_mps[!all_mps %in% unique(ora_results_df$metaprogram)] |>
  paste0(collapse = ", ")

if(missing_mps != "") {
  glue::glue("
      <div class=\"alert alert-info\">

      No gene sets were identified as significant in the following metaprograms:
      {missing_mps}
      </div>
    ")
  }
```


To evaluate if the MPs are easily interpretable, we will make a heatmap that shows the significance of each gene set across all MPs.
Ideal MPs will show gene sets that are uniquely enriched in that MP instead of enriched across multiple MPs.
The rows represent metaprograms, the columns represent all significant gene sets, and the heatmap color indicates the -log10(adjusted p-value) for that gene set in that metaprogram.
Only the top 5 gene sets for each metaprogram are shown to save space.

Only MPs with identified gene sets will be included.

```{r}
#| fig.height: 10
#| fig.width: 22


# convert ora results into a heatmap of pvalues
heatmap_mtx <- ora_results_df |>
  dplyr::mutate(logp = -log10(p.adjust)) |>
  # only keep the top 5 gene sets
  dplyr::slice_max(logp, n = 5, by = metaprogram) |>
  dplyr::select(metaprogram, ID, logp) |>
  tidyr::pivot_wider(
    names_from = metaprogram,
    values_from = logp,
    values_fill = 0  # set missing gene sets to 0
  ) |>
  # remove _ and wrap gene set names
  dplyr::mutate(
    ID = stringr::str_replace_all(ID, "_", " ") |>
      stringr::str_wrap(width = 33)
  ) |>
  tibble::column_to_rownames("ID") |>
  as.matrix() |>
  t()

ComplexHeatmap::Heatmap(
  heatmap_mtx,
  name = "-log10(adjusted p-value)\n",
  row_names_side = "left",
  row_dend_side = "right",
  cluster_rows = FALSE,
  col = circlize::colorRamp2(c(0, max(heatmap_mtx)), colors = c("white", "darkslateblue")),
  row_names_gp = grid::gpar(fontsize = 20),
  column_names_gp = grid::gpar(fontsize = 16),
  heatmap_legend_param = list(
    title_gp = grid::gpar(fontsize = 20),
    labels_gp = grid::gpar(fontsize = 20)
  )
) |>
  # don't cut off super long gene set names
  ComplexHeatmap::draw(padding = unit(c(6, 2, 2, 2), "cm"))
```

Alternatively, we can use the Jaccard similarity to index to measure how similar the list of gene sets for each MP is to all other MPs.


```{r}
ora_jaccard_heatmap(ora_results_df,
                    title = "All gene sets")
```


### Specificity and similarity of gene sets

Here we will look at the identified gene sets for each metaprogram.
If multiple gene sets are returned, it's likely that they are biologically similar.
Having multiple gene sets that are biologically distinct is going to impede interpretation.

```{r}
#| fig.height: 25
#| fig.width: 15

# dot plots for all mps with at least 1 significant gene set
split(ora_results_df, ora_results_df$metaprogram) |>
  purrr::imap(\(result, mp){
    if(nrow(result) > 0) {
      dotplot(result, title = mp)
    }
  }) |>
  purrr::compact() |>
  patchwork::wrap_plots(ncol = 2)
```


## Evaluating metaprogram scores across individual cells

Good MPs should show variation in scores for MPs across samples.

Score were assigned for each cell by getting the dot product between the weights for the top genes in the MP and the normalized gene expression values.
Before calculating the score, gene expression was quantile normalized within each sample.

First we look at the score distribution for all samples for each metaprogram.
Good MPs will show variability within the dataset, in particular we are looking for a bimodal distribution.

```{r}
#| message: FALSE
#| warning: FALSE

ggplot(score_df,
       aes(x = mp_score, y = forcats::fct_rev(metaprogram))) +
  ggridges::geom_density_ridges(alpha = 0.8, color = "white", rel_min_height = 0.01) +
  theme(
    strip.background = element_rect(fill = "grey90"),
    axis.text.y = element_text(size = 8),
    legend.position = "none"
  ) +
  labs(
    x = "Score across all samples",
    y = "Metaprogram"
  )
```

Good metaprograms should show variation in the scores assigned to cells in the original dataset.
To measure variation, we will use the coefficient of variation (CV), which reports a score's variance relative to the mean score across all cells.

```{r}
# calculate the variance for each MP in each sample
mp_var_df <- score_df |>
  dplyr::summarise(
    variance = stats::var(mp_score, na.rm = TRUE),
    cv = sqrt(variance) / mean(mp_score),
    .by = c("sample_id", "metaprogram")
  )

ggplot(
  mp_var_df,
  aes(
    x = metaprogram,
    y = cv
  )
) +
  ggforce::geom_sina(size = 1, alpha = 0.8) +
  labs(
    x = "Metaprogram",
    y = "CV across all cells in a sample"
  )

```

Now, we'll look at the distribution for each library individually.
Good MPs should show variability in multiple samples.

```{r}
# dynamic figure height for the per sample plot below, which facets with 4 columns, so the
# height scales with the number of rows needed to fit one panel per sample
n_samples <- dplyr::n_distinct(score_df$sample_id)
score_facet_ncol <- 4
dynamic_score_fig_height <- ceiling(n_samples / score_facet_ncol) * 3
```

```{r}
#| fig.height: !r dynamic_score_fig_height
#| fig.width: 12
#| message: FALSE
#| warning: FALSE


ggplot(score_df,
       aes(x = mp_score, y = metaprogram, fill = metaprogram)) +
  ggridges::geom_density_ridges(alpha = 0.8, color = "white", bw = 0.5, rel_min_height = 0.1) +
  facet_wrap(vars(sample_id), scales = "free_x", ncol = score_facet_ncol) +
  theme(
    strip.background = element_rect(fill = "grey90"),
    axis.text.y = element_text(size = 6),
    legend.position = "none"
  ) +
  labs(
    x = "Score",
    y = "Metaprogram",
    title = "MP scores by sample"
  )
```


## Session info

```{r}
sessionInfo()
```
