---
title: "Comparing metaprograms generated using a range of k"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    code_folding: hide
params:
  metaprogram_metrics_file: NULL
  k_metrics_file: NULL
  neff_background_file: NULL
  specificity_background_file: NULL
  cv_background_file: NULL
  optimal_k_file: NULL
  output_heatmap_file: NULL
  seed: 2026
---

<!--TODO: Update this report and move it to a separate process.
Right now this is not used in the workflow.
-->

This report compares metaprograms generated from a single cohort across a range of k values.
Here k refers to the number of clusters specified during hierarchical clustering of spectra to define metaprograms.

All metrics shown in this report, and the optimal value of k, are calculated by `calculate-optimal-k.R`.
This report only reads in the tables written by that script and uses them to make plots.

## Set Up

```{r}
# helper function for plotting permutation results
# plots background points with observed values in color on top as sina plot
plot_permutation_results <- function(background_df, observed_df, x_axis, y_axis){

  ggplot(background_df, aes(x = {{ x_axis }}, y = {{ y_axis }})) +
    ggforce::geom_sina(
      color = "gray",
      alpha = 0.1
    ) +
    geom_point(
      data = observed_df,
      aes(
        x = {{ x_axis }},
        y = {{ y_axis }},
        color = is_significant
      ),
      size  = 3,
      shape = 18  # diamond
    ) +
    scale_color_manual(values = c(`TRUE` = "red", `FALSE` = "blue")) +
    theme(
      axis.text.x = element_text(angle = 90)
    )

}
```

```{r setup, message = FALSE, echo = FALSE}
library(ggplot2)

# Set default ggplot theme
theme_set(
  theme_bw() +
    theme(
      strip.background = element_rect(fill = "transparent", linewidth = 0.5),
      strip.text = element_text(face = "bold"),
      axis.text.x = element_text(angle = 90)
    )
)

options(DT.options = list(
  scrollX = TRUE,
  scrollY = "400px",
  scrollCollapse = TRUE,
  paging = FALSE,
  searching = FALSE
))

# the sina and beeswarm plots are jittered, so set a seed to keep the report reproducible
set.seed(params$seed)

# cutoffs applied by `calculate-optimal-k.R`
# they are repeated here so they can be shown on the plots and must match the values used there
sig_value <- 0.05
correlation_cutoff <- 0.5
coherence_cutoff <- 0.1
```

```{r}
# make sure all inputs were provided and exist
# a list is used rather than a vector so that any missing params are still present to check
input_files <- list(
  params$metaprogram_metrics_file,
  params$k_metrics_file,
  params$neff_background_file,
  params$specificity_background_file,
  params$cv_background_file,
  params$optimal_k_file
)

stopifnot(
  "All input files from the calculate optimal k script must be provided" = !any(sapply(input_files, is.null)),
  "Some or all input files do not exist" = all(file.exists(unlist(input_files))),
  "Output heatmap file must be provided" = !is.null(params$output_heatmap_file)
)
```

```{r}
# metrics for each metaprogram within each value of k
# only values of k included in the comparison are present here
mp_metrics_df <- readr::read_tsv(params$metaprogram_metrics_file, show_col_types = FALSE)

# summarized metrics and the normalized rank of each metric for each value of k
# any values of k that were excluded are included here with no metrics
k_metrics_df <- readr::read_tsv(params$k_metrics_file, show_col_types = FALSE)

# null distributions used to assess significance of the observed metrics
neff_background_df <- readr::read_tsv(params$neff_background_file, show_col_types = FALSE)
specificity_background_df <- readr::read_tsv(params$specificity_background_file, show_col_types = FALSE)
cv_background_df <- readr::read_tsv(params$cv_background_file, show_col_types = FALSE)

# the optimal value of k, reported as a number
optimal_k <- readr::read_lines(params$optimal_k_file) |>
  as.numeric()

# the tables label each value of k as `k_05`, so match that formatting to pull out the optimal k
optimal_k_label <- sprintf("k_%02d", optimal_k)

stopifnot(
  "The optimal value of k does not match the value of k flagged in the k metrics table" =
    identical(
      optimal_k_label,
      k_metrics_df |> dplyr::filter(is_optimal_k) |> dplyr::pull(n_metaprograms)
    )
)
```

```{r}
# set figure heights based on if we have the smaller set of ks (5-15)
# or the larger set of ks (5-25)
total_k <- dplyr::n_distinct(mp_metrics_df$n_metaprograms)
if (total_k <= 12) {
  dynamic_fig_height <- 10
} else {
  dynamic_fig_height <- 20
}
```

```{r}
# report any values of k that were removed before calculating metrics
excluded_k <- k_metrics_df |>
  dplyr::filter(is_excluded) |>
  dplyr::pull(n_metaprograms)

knitr::asis_output(
  glue::glue("The following values of k have at least one metaprogram with only one spectra and were removed from this analysis:
             {paste0(excluded_k, collapse = ', ')}
             "
  )
)
```

```{r}
# account for the fact that sample types might not be present
# use this to skip any chunks that are sample type specific
has_sample_types <- all(c("num_primary_samples", "num_mets_samples") %in% colnames(mp_metrics_df))
```


## Sample diversity

Metaprograms should be representative of a diverse set of samples.
This section will summarize the proportion of samples found in each metaprogram.

```{r}
# display total samples
# this is the number of unique ids that contribute cell scores
total_samples <- max(k_metrics_df$total_samples, na.rm = TRUE)

knitr::asis_output(
  glue::glue("The metaprograms explored here are comprised of **{total_samples} samples**.")
)
```


### Proportion of samples

Here each dot is a metaprogram and the value indicated is the proportion of total **samples** represented by spectra in that metaprogram.
The dotted line indicates 10% which may help identify any metaprograms with a low representation of samples.

```{r}
ggplot(mp_metrics_df, aes(x = n_metaprograms, y = proportion_of_samples)) +
  geom_boxplot() +
  ggbeeswarm::geom_beeswarm() +
  labs(
    x = "Number of metaprograms (k)",
    y = "Proportion of total samples \n contributing to each metaprogram"
  )  +
  geom_hline(yintercept = 0.1, linetype = "dashed")
```


### Representation across sample types

Here we will look at the contribution of different sample types to each of the metaprograms.
Particularly, we will see how many of the samples are from primary or metastatic sites within each metaprogram.
If this information is not present, no plots or tables will be shown.

```{r}
#| eval: !r has_sample_types

sample_type_df <- mp_metrics_df |>
  tidyr::pivot_longer(
    cols = c("num_primary_samples", "num_mets_samples"),
    names_to = "sample_type",
    values_to = "count"
  ) |>
  dplyr::mutate(
    proportion = count/ num_samples,
    sample_type = stringr::word(sample_type, 2, sep = "_") |>
      factor(labels = c("Primary", "Metastasis"))
  ) |>
  dplyr::select(n_metaprograms, metaprogram, proportion, sample_type)
```

```{r}
#| fig.height: !r dynamic_fig_height
#| fig.width: 15
#| eval: !r has_sample_types

ggplot(sample_type_df, aes(x = metaprogram, y = proportion, fill = sample_type)) +
  geom_col(width = 0.7) +
  facet_wrap(vars(n_metaprograms), scales = "free_x", ncol = 4) +
  scale_fill_manual(values = c(Primary = "#6A9FC0", Metastasis = "#C47A8A")) +
  scale_y_continuous(labels = scales::percent_format(), expand = expansion(mult = c(0, 0.02))) +
  labs(
    x    = "Metaprogram",
    y    = "Proportion of samples",
    fill = "Sample type"
  ) +
  theme(
    strip.text         = element_text(face = "bold"),
    legend.position    = "top",
    axis.text.x        = element_text(angle = 90, hjust = 1, size = 10)
  )
```

### Normalized effective sample size

Here we will look at the normalized effective species size or Hill number of order 1.
This is calculated by taking the exponential of Shannon's entropy and represents the effective number of common or abundant samples.
Here we are looking for a number close to the total possible samples in that metaprogram.
The values are normalized by the total number of samples in the metaprogram such that values close to 1 indicate all samples are equally abundant.
Note that every unique ID in the input samplesheet is treated as a separate sample.

To determine if the observed effective sample size is not due to random chance, we performed permutation testing.
The spectra assigned to each metaprogram were shuffled while preserving the total number of spectra assigned to each metaprogram.
After shuffling, the normalized effective sample size was calculated to create a null distribution.

The observed value was compared to the null distribution to get an empirical p-value, producing a single p-value for each metaprogram within a value of k.
The below plot shows both the null distribution in gray and the observed value for each metaprogram.
The observed values are colored in red if they are significantly greater than the null distribution.


```{r}
#| fig.height: !r dynamic_fig_height
#| fig.width: 15

# the plotting function colors points by `is_significant`
neff_obs_df <- mp_metrics_df |>
  dplyr::select(n_metaprograms, metaprogram, eff_number_norm, is_significant = neff_is_significant)

plot_permutation_results(neff_background_df, neff_obs_df, metaprogram, eff_number_norm) +
  facet_wrap(vars(n_metaprograms), scale = "free_x", ncol = 3) +
  labs(
    y = "Normalized effective sample size",
    x = "Metaprogram",
    color = "significant"
  ) +
  theme(
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 14)
  )
```

We then can identify the percentage of metaprograms for each value of k that have a significant normalized effective sample size.
This is used to identify the optimal value of k in the final section.

```{r}
k_metrics_df |>
  dplyr::filter(!is_excluded) |>
  dplyr::select(n_metaprograms, neff_sig_number, neff_sig_proportion) |>
  DT::datatable(colnames = c("k", "Number of metaprograms with significant effective sample size", "% of metaprograms with significant effective sample size")) |>
  DT::formatPercentage("neff_sig_proportion", 2)
```

## Correlation

Metaprograms should be unique and non-redundant.
Therefore we expect them to have low correlation to other metaprograms from the same value of k.

Here the pairwise correlation of gene weights across all metaprograms was calculated.
The plot then shows the maximum correlation value for each metaprogram.
Metaprograms with a max correlation of > `r correlation_cutoff` are considered redundant.
The number of metaprograms above this cutoff is used to determine the optimal value of k.


```{r}
ggplot(mp_metrics_df, aes(x = n_metaprograms, y = max_correlation)) +
  geom_boxplot(outlier.shape = NA) +
  ggbeeswarm::geom_beeswarm(aes(color = correlation_is_passing)) +
  scale_color_manual(values = c(`FALSE` = "red", `TRUE` = "black")) +
  labs(
    x = "Number of metaprograms (k)",
    y = "Max correlation of metaprogram gene weights",
    color = glue::glue("Max correlation ≤ {correlation_cutoff}")
  )
```

```{r}
k_metrics_df |>
  dplyr::filter(!is_excluded) |>
  dplyr::select(n_metaprograms, num_passing_correlation, proportion_passing_correlation) |>
  DT::datatable(colnames = c("k", glue::glue("Number of metaprograms with max correlation ≤ {correlation_cutoff}"), glue::glue("% of metaprograms with max correlation ≤ {correlation_cutoff}"))) |>
  DT::formatPercentage("proportion_passing_correlation", 2)
```


## Coherence

Metaprograms should show coherence, meaning the spectra that comprise the metaprogram are more similar to other spectra in the same metaprogram than to spectra in different metaprograms.
To measure this, we calculated the correlation of each spectra to all other spectra.
We then calculate the difference in medians between the distribution of correlation of spectra within a given metaprogram and between metaprograms, indicated as "coherence" throughout this report.

Metaprograms with coherence <= `r coherence_cutoff` are likely to represent noise and will not be used in further analysis.

```{r}
ggplot(mp_metrics_df, aes(x = n_metaprograms, y = median_difference)) +
  geom_boxplot() +
  ggbeeswarm::geom_beeswarm(aes(color = coherence_is_passing)) +
  scale_color_manual(values = c("red", "black")) +
  labs(
    x = "Number of metaprograms (k)",
    y = "Metaprogram coherence",
    color = glue::glue("Coherence > {coherence_cutoff}")
  )

```

This plot suggests that increasing k doesn't improve metaprogram coherence after certain values of k.
We therefore make an elbow plot showing the median coherence for each value of k and find the elbow.

```{r}
# the median coherence for each value of k, only for the values of k that were compared
coherence_elbow_df <- k_metrics_df |>
  dplyr::filter(!is_excluded)

ggplot(coherence_elbow_df, aes(x = n_metaprograms, median_coherence)) +
  geom_point() +
  labs(
    x = "Number of metaprograms (k)",
    y = "Median coherence"
  )
```

The elbow is identified with a modified version of the [kneedle algorithm](https://raghavan.usc.edu/papers/kneedle-simplex11.pdf).
Here we only find one global maximum rather than a set of local maxima.
Values of k greater than the elbow indicate no substantial gain of coherence over the lower values of k.

```{r}
elbow_value <- coherence_elbow_df |>
  dplyr::filter(is_coherence_elbow) |>
  dplyr::pull(k)

knitr::asis_output(
  glue::glue("**Value of k corresponding to the elbow: {elbow_value}**")
)
```

## Gene set specificity

After generating metaprograms for each value of k, the top 200 genes based on gene weights were extracted from each metaprogram.
Over-representation analysis was then run using `clusterProfiler::enricher()` function on these 200 genes.
Input gene sets include a set of custom `EWS::FLI1` activated and repressed gene sets and Hallmarks from `MSigDb`.

Any gene sets with an adjusted p value ≤ `r sig_value` are considered significant.
Good metaprograms should have at least one significant gene set.

The table below shows the total number of metaprograms with at least one significant gene set for each value of k.

```{r}
k_metrics_df |>
  dplyr::filter(!is_excluded) |>
  dplyr::select(n_metaprograms, num_metaprograms_with_genesets, proportion_metaprograms_with_genesets) |>
  DT::datatable(colnames = c("k", "Number of metaprograms with ≥ 1 significant gene set", "% of metaprograms with ≥ 1 significant gene set")) |>
  DT::formatPercentage("proportion_metaprograms_with_genesets", 2)
```


Metaprograms should be biologically distinct and annotated by a unique set of gene sets.
This means gene sets should be significant in as few metaprograms as possible, ideally only one metaprogram.
For each gene set, we calculated gene set specificity using the following formula:

$$1 - \frac{\text{# of metaprograms the gene set is significant in}- 1}{\text{total # of metaprograms} - 1}$$

A score of 1 indicates the gene set is unique to a single metaprogram, while a score of 0 indicates it is shared across all metaprograms.

We then take the mean specificity across all gene sets and all metaprograms for a single value of k.
Metaprograms with no identified gene sets are not considered in this calculation.

To determine if this value is greater than what would be observed by random chance, we generated a null distribution by shuffling the gene sets assigned to each metaprogram, keeping the number of gene sets consistent with the observed distribution and recalculating the mean specificity.
When shuffling the gene sets, each gene set was weighted based on the number of metaprograms in which the gene set was considered significant.

The below plot shows the observed mean gene set specificity and the associated background distribution for each value of k.
Any values of k with a significant mean gene set specificity are indicated in red.

```{r}
gss_obs_df <- k_metrics_df |>
  dplyr::filter(!is_excluded) |>
  dplyr::select(n_metaprograms, overall_geneset_specificity, geneset_specificity_significant) |>
  dplyr::mutate(
    is_significant = geneset_specificity_significant == 1
  )

plot_permutation_results(specificity_background_df, gss_obs_df, n_metaprograms, overall_geneset_specificity) +
  labs(
    y = "Mean gene set specificity",
    x = "Number of metaprograms (k)",
    color = "significant"
  )
```


## Variation in cell scores

The gene weights for the top 200 genes are used to calculate a cell score for each cell in the dataset.
This is done by taking the dot product of the gene weights and quantile-normalized counts in each cell.

Good metaprograms should show variation in the scores assigned to cells in the original dataset.
To measure variation, we will use the coefficient of variation (CV), which reports a score's variance relative to the mean score across all cells.

Here we compare the observed CV to a null distribution to determine if a metaprogram shows significantly more variation compared to random chance.
The background distribution was constructed by shuffling the gene identifiers for each metaprogram in the gene by metaprogram matrix.
This ensured that both the weights for each metaprogram and the relationship of genes across metaprograms were retained.
After shuffling, the top 200 genes were extracted, and the scores were re-calculated.
This produces a unique background distribution for each metaprogram.

```{r}
#| fig.height: !r dynamic_fig_height
#| fig.width: 15

cv_obs_df <- mp_metrics_df |>
  dplyr::select(n_metaprograms, metaprogram, score_cv, is_significant = cv_is_significant)

plot_permutation_results(cv_background_df, cv_obs_df, metaprogram, score_cv) +
  facet_wrap(
    vars(n_metaprograms),
    scales = "free_x",
    ncol = 4
  ) +
  labs(
    x = "Metaprogram",
    y = "Coefficient of variation (CV)",
    color = "Significant"
  ) +
  theme(
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 14)
  )
```

The below table shows the total percentage of metaprograms with significant CV for each value of k.

```{r}
k_metrics_df |>
  dplyr::filter(!is_excluded) |>
  dplyr::select(n_metaprograms, num_significant_cv, proportion_significant_cv) |>
  DT::datatable(colnames = c("k", "Number of metaprograms with significant CV", "% of metaprograms with significant CV")) |>
  DT::formatPercentage("proportion_significant_cv", 2)
```

## Combining metrics to identify an optimal value of k

The above metrics are combined to define the optimal value of k.
Each metric has a single number reported for each value of k on a 0-1 scale where 1 indicates a good metaprogram.
The values included are:

* `Effective sample size`: Proportion of metaprograms with a normalized effective sample size significantly greater than the null distribution (p ≤ `r sig_value`).
* `Correlation`: Proportion of metaprograms with a max correlation to all other metaprograms < `r correlation_cutoff`.
* `Coherence`: Value indicating if the specific k is at, above, or below the elbow point for coherence.
The value of k at the elbow is prioritized and assigned a value of 1.
Values of k below the elbow will be 0.5 and values above the elbow will be 0.
* `Gene set specificity`: Binary value indicating if the overall gene set specificity is significantly greater than the null distribution (p ≤ `r sig_value`).
Values of k that are significant will be 1 and those that are not significant will be 0.
* `Coefficient of variation`: Proportion of metaprograms with a CV significantly greater than the null distribution (p ≤ `r sig_value`).

For each metric, each value of k is ranked, assigning the same rank to any values of k with the exact same value for a specific metric.
The rank is normalized by the total number of possible k so that the scale is still 0-1, where 1 indicates a good metaprogram.
The optimal value of k is the k with the highest average rank.

```{r}
# the rank columns of the k metrics table, in the order they are shown in the heatmap
# the names are used as the row labels, so they are defined here with the rest of the plotting code
rank_columns <- c(
  "Effective sample size" = "rank_effective_sample_size",
  "Correlation" = "rank_correlation",
  "Coherence" = "rank_coherence",
  "Gene set specificity" = "rank_gene_set_specificity",
  "Coefficient of variation" = "rank_coefficient_of_variation",
  "Mean rank" = "mean_rank"
)

# selecting with the named vector labels each row of the heatmap, with one column per value of k
rank_mtx <- k_metrics_df |>
  dplyr::filter(!is_excluded) |>
  dplyr::select(n_metaprograms, dplyr::all_of(rank_columns)) |>
  tibble::column_to_rownames("n_metaprograms") |>
  as.matrix() |>
  t()
```

```{r}
# first export the heatmap
pdf(params$output_heatmap_file, width = 10, height = 5)

# define a color scale
col_fun <- circlize::colorRamp2(
  breaks = c(0, 0.75, 1),
  colors = c("#2166AC", "white", "#B2182B")
)

# make the heatmap
ht <- ComplexHeatmap::Heatmap(
  rank_mtx,
  name = "Value",
  col = col_fun,
  # don't cluster and preserve our ordering
  cluster_rows    = FALSE,
  cluster_columns = FALSE,
  # Split last row into its own section
  row_split = c(rep("Metrics", nrow(rank_mtx) - 1), "Mean rank"),
  row_gap   = unit(0.3, "cm"),  # gap size between the two sections
  # Labels
  row_names_side    = "left",
  column_names_side = "bottom",
  row_names_gp      = grid::gpar(fontsize = 10),
  column_names_gp   = grid::gpar(fontsize = 10),
  column_names_rot  = 90,
  # Cell appearance
  rect_gp = grid::gpar(col = "white", lwd = 2), # white cell border
  border  = FALSE, # no border around the whole thing
  # Legend
  heatmap_legend_param = list(
    title          = "Value",
    title_position = "leftcenter-rot", # rotate title on left
    legend_height  = unit(4, "cm"),
    at             = c(0, 0.25, 0.5, 0.75, 1),
    labels         = c("0", "0.25", "0.5", "0.75", "1")
  ),
  # Axis labels
  row_title      = "Metric",
  row_title_side = "left",
  row_title_gp   = grid::gpar(fontsize = 11),
  # Square cells: set a fixed size per cell
  cell_fun = NULL,
  width  = unit(ncol(rank_mtx) * 0.8, "cm"),  # 0.8 cm per column
  height = unit(nrow(rank_mtx) * 0.8, "cm")   # 0.8 cm per row (same as width = square)
)

ComplexHeatmap::draw(ht, heatmap_legend_side = "right")

dev.off()
```

```{r}
#| fig.width: 10

# now render the heatmap for the report
ComplexHeatmap::draw(ht, heatmap_legend_side = "right")
```


```{r}
# find any metaprograms in the optimal k that have low coherence
low_coherence_mps <- mp_metrics_df |>
  dplyr::filter(!coherence_is_passing, n_metaprograms == optimal_k_label) |>
  dplyr::pull(metaprogram)
```

```{r}
# compile text based on if there are any metaprograms with low coherence
main_text <- glue::glue(
  "**The optimal value of k for this cohort is: {optimal_k}**"
)

coherence_text <- glue::glue(
  "For this value of k, {length(low_coherence_mps)} metaprogram(s) do not pass the minimum coherence of {coherence_cutoff}. "
)

if (length(low_coherence_mps) > 0) {
  # paste the metaprograms into a single string so the line isn't repeated for each metaprogram
  coherence_text <- glue::glue(
    "{coherence_text}
    This includes {paste0(low_coherence_mps, collapse = ', ')}."
  )
}

knitr::asis_output(
  glue::glue(
    "
    {main_text}

    {coherence_text}
    "
  )
)
```


## Session info

```{r}
sessionInfo()
```
